Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2024-12-20 ms-3/ms.tex
Update manuscript and format some code

\clearpage
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses. 
However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data. We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.

@@

Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers (i.e., the expected number of secondary infections that will result if a particular individual becomes infected), quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}.
However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\citep{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\citep{sun2021transmission}. 
An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.

@@

What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
Under a wide range of configurations for a branching process model the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter. 
The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. 
Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial. 
Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\citep{barron1992analysis}.

@@

In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, incidence is often modeled as
\begin{equation}
    I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
\end{equation}
where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}. 
Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.

@@

Modeling the underlying epidemic dynamics based on log-transformed incidence allows us to address the statistical effects of population size on the relationship between the mean and variance in count data, which would otherwise confound our analysis.
Specifically, since population size influences the mean and variance of case count data, it impacts estimates of dispersion in populations of different sizes that are otherwise identical.
Accordingly, population size appears as an offset in our model of broad-scale incidence changes. That is,
\begin{equation}
  \log(\mu_t) = \sum_{j=1}^J \beta_j^{(t)} h_j(t) + \log(N) 
\end{equation}

@@

Since the 'glm.nb' function in the MASS R package \cite{MASS} doesn't accomodate an offset term, we estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$. That is, for each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ was estimated using the spline and offset term via IRLS in the NBPSeq R package~\citep{di2015nbpseq}. Then, a single value of $\theta_t$ for the time window was computed via an optimization procedure.

In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that could be applied at each time step to test the hypothesis that $\theta$ has changed. This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model. For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the \begin{math}\theta\end{math}-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.

@@

To test the validity and power of the methods, we epidemic curves with known values of $\mu$ and $\theta$.
We varyied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
Epidemic curves were created from a deterministic skeleton (template) that either assumed a flat distribution of incidence over time or a Gaussian-shaped distribution, with parameters controlling the total epidemic size, peak time, peak width, and duration. 
Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $theta_2$ were used in each respective piece, before and after the breakpoint. 
This allowed us to simulate epidemic curves with variation in dispersion, as well as the special case when $\theta_1 = \theta_2$.

\subsection*{Data availability and processing}
We applied our method to cumulative COVID-19 case countes in US counties~\citep{nytimes_covid19}.
Population sizes were from The United States Census Bureau~\citep{us_census_county_population_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases. Missing values for new cases were all at the beginning of the pandemic and were imputed as zero. 
Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero.

>>>>

\clearpage
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data. We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.

@@

Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}, indicating temporal changes in the level of variability around the expected number of secondary infections that will result if a particular individual becomes infected.
However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\citep{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\citep{sun2021transmission}. 
An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.

@@

What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
Under a wide range of configurations for a branching process model of contagion spread the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter. 
The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. 
Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial. 
Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\citep{barron1992analysis}.

@@

In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, the number of infections each generation is often modeled as
\begin{equation}
    I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
\end{equation}
where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}. 
Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.

@@

Modeling the underlying epidemic dynamics based on log-transformed incidence allows us to address the statistical effects of population size on the relationship between the mean and variance in count data, which would otherwise confound our analysis.
Specifically, since population size influences the mean and variance of case count data, it impacts dispersion in different-sized populations that are otherwise identical.
Accordingly, population size appears as an offset in our model of broad-scale incidence changes. That is,
\begin{equation}
  \log(\mu_t) = \sum_{j=1}^J \beta_j^{(t)} h_j(t) + \log(N) 
\end{equation}

@@

We estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$. That is, for each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ was estimated using the spline and offset term via IRLS in the NBPSeq R package~\citep{di2015nbpseq}. Then, a single value of $\theta_t$ for the time window was computed via an optimization procedure.

In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that could be applied at each time step to test the hypothesis that $\theta$ has changed. This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part ($\theta$-change) model. For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the $\theta$-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.

@@

To test the validity and power of the methods, we epidemic curves with known values of $\mu$ and $\theta$.
We varied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
Epidemic curves were created from a deterministic skeleton (template) that either assumed a flat distribution of incidence over time or a Gaussian-shaped distribution, with parameters controlling the total epidemic size, peak time, peak width, and duration. 
Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $\theta_2$ were used in each respective piece, before and after the breakpoint. 
This allowed us to simulate epidemic curves with variation in dispersion, as well as the special case when $\theta_1 = \theta_2$.

\subsection*{Data availability and processing}
We applied our method to cumulative COVID-19 case countes in US counties~\citep{nytimes_covid19}.
Population sizes were from The United States Census Bureau~\citep{us_census_county_population_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases. Missing values for new cases were all at the beginning of the pandemic and were imputed as zero. 
Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
We used a width paramter of $\delta = 8$ and set the degrees of freedom parameter for the natural splines to $J = 3$.
